---
title: "Session 8: Learning New Tools with AI (DuckDB)"
execute:
    eval: false
---

## Introduction

In this session, you will learn a powerful new tool — **DuckDB** — with AI as your guide. The goal is not to be spoon-fed answers, but to use AI to **understand** how the tool works so you can apply it independently.

::: callout-important
## Upskilling, not de-skilling
We will use AI to explain concepts and generate starter code, but you must **read, understand, and question** every output. If you don't understand what the code does, ask the AI to explain it. The aim is to leave this session *more capable* than when you started.
:::

## What is DuckDB?

DuckDB is an **in-process SQL database** — think of it as "SQLite for analytics." Key features:

-   Runs inside R or Python (no server needed)
-   Extremely fast for analytical queries on large datasets
-   Can query CSV, Parquet, and GeoJSON files directly
-   Supports spatial data via the `spatial` extension

### Why DuckDB for Transport?

Transport analysis often involves:

-   Large datasets (millions of OSM road segments, GTFS records, GPS traces)
-   Complex filtering and aggregation (group by zone, time period, mode)
-   Joining multiple data sources (OD data + network data + zone boundaries)

DuckDB excels at these tasks, often outperforming R's dplyr or Python's pandas for large data.

## Part 1: Learning DuckDB with AI (45 mins)

### Step 1: Ask AI to Explain DuckDB

Start a conversation with Copilot Chat:

```
Explain DuckDB to a transport planner who knows R/Python but has never used 
a SQL database. What makes it different from data.frame/pandas? 
Give 3 concrete examples relevant to transport data analysis.
```

Read the response carefully. Ask follow-up questions:

-   "How do I load DuckDB in R vs Python?"
-   "What's the difference between an in-memory and file-based database?"
-   "When would I use DuckDB instead of dplyr/pandas?"

### Step 2: Your First DuckDB Query

Ask Copilot:

```
Write R code to create a DuckDB database, load leeds_network.geojson into it, 
and run a SQL query that counts the number of road segments by highway type, 
sorted by count descending. Explain each step.
```

::: {.panel-tabset}

### R

```{r}
#| label: duckdb-first-r
#| eval: false
# Try this yourself first, then compare with AI output!
library(duckdb)
library(sf)

# Connect to DuckDB (in-memory)
con = dbConnect(duckdb())

# Install and load spatial extension
dbExecute(con, "INSTALL spatial")
dbExecute(con, "LOAD spatial")

# Load the Leeds network data
dbExecute(con, "
  CREATE TABLE leeds_network AS 
  SELECT * FROM st_read('leeds_network.geojson')
")

# Query: count segments by highway type
result = dbGetQuery(con, "
  SELECT highway, COUNT(*) as n_segments
  FROM leeds_network
  GROUP BY highway
  ORDER BY n_segments DESC
")
print(result)
```

### Python

```{python}
#| label: duckdb-first-py
#| eval: false
# Try this yourself first, then compare with AI output!
import duckdb

# Connect to DuckDB (in-memory)
con = duckdb.connect()

# Install and load spatial extension
con.execute("INSTALL spatial")
con.execute("LOAD spatial")

# Load the Leeds network data
con.execute("""
  CREATE TABLE leeds_network AS 
  SELECT * FROM st_read('leeds_network.geojson')
""")

# Query: count segments by highway type
result = con.execute("""
  SELECT highway, COUNT(*) as n_segments
  FROM leeds_network
  GROUP BY highway
  ORDER BY n_segments DESC
""").df()
print(result)
```

:::

::: {.callout-tip}
## Check your understanding
Before moving on, make sure you can answer: What does `GROUP BY` do? Why is `ORDER BY n_segments DESC` useful? If unsure, ask the AI!
:::

## Part 2: OD Data with DuckDB (45 mins)

Now we bring in origin-destination data. The R `od` package provides example datasets that work well for learning.

### Step 3: Loading OD Data

Ask Copilot:

```
The R 'od' package has example OD datasets. Write R code to:
1. Load the od::od_leeds dataset
2. Save it as a CSV
3. Load it into DuckDB
4. Query the top 10 OD pairs by bicycle count

Also show the Python equivalent using DuckDB directly on the CSV.
```

::: {.panel-tabset}

### R

```{r}
#| label: duckdb-od-r
#| eval: false
library(duckdb)
library(od)
library(dplyr)

# Get Leeds OD data
od_data = od::od_leeds
# Save for DuckDB
write.csv(od_data, "leeds_od_data.csv", row.names = FALSE)

# Connect to DuckDB
con = dbConnect(duckdb())
dbExecute(con, "LOAD spatial")

# Load OD data
dbExecute(con, "
  CREATE TABLE od_data AS 
  SELECT * FROM read_csv_auto('leeds_od_data.csv')
")

# Top 10 cycling flows
top_cycle = dbGetQuery(con, "
  SELECT * FROM od_data
  ORDER BY bicycle DESC
  LIMIT 10
")
print(top_cycle)

# Total trips by mode
mode_totals = dbGetQuery(con, "
  SELECT 
    SUM(bicycle) as total_bicycle,
    SUM(walk) as total_walk,
    SUM(car_driver) as total_car,
    SUM(all) as total_all
  FROM od_data
")
print(mode_totals)
```

### Python

```{python}
#| label: duckdb-od-py
#| eval: false
import duckdb

# Connect to DuckDB
con = duckdb.connect()
con.execute("LOAD spatial")

# Load OD data directly from CSV
con.execute("""
  CREATE TABLE od_data AS 
  SELECT * FROM read_csv_auto('leeds_od_data.csv')
""")

# Top 10 cycling flows
top_cycle = con.execute("""
  SELECT * FROM od_data
  ORDER BY bicycle DESC
  LIMIT 10
""").df()
print(top_cycle)

# Mode share analysis
mode_totals = con.execute("""
  SELECT 
    SUM(bicycle) as total_bicycle,
    SUM(walk) as total_walk,
    SUM(car_driver) as total_car,
    SUM(all) as total_all
  FROM od_data
""").df()
print(mode_totals)
```

:::

### Step 4: Spatial Queries with DuckDB

Ask Copilot:

```
Write code to join the Leeds OD data with zone boundaries using DuckDB's 
spatial extension. Calculate the total cycling trips originating from each zone 
and create a choropleth map of the results.
```

::: {.callout-note}
## Challenge
Can you write this query *without* looking at the AI answer first? Try it, then compare.
:::

## Part 3: Routing Data with DuckDB (30 mins)

### Step 5: Network Analysis

Ask Copilot:

```
Using the leeds_network.geojson loaded in DuckDB, write a query that:
1. Calculates the total length of cycleway in metres
2. Finds the average length of road segments by highway type
3. Identifies the longest continuous road (by name)

Explain the SQL functions used.
```

::: {.panel-tabset}

### R

```{r}
#| label: duckdb-network-r
#| eval: false
# Total cycleway length
cycleway_stats = dbGetQuery(con, "
  SELECT 
    highway,
    COUNT(*) as n_segments,
    SUM(ST_Length(geom)) as total_length_m,
    AVG(ST_Length(geom)) as avg_segment_length_m
  FROM leeds_network
  WHERE highway IN ('cycleway', 'path', 'footway')
  GROUP BY highway
  ORDER BY total_length_m DESC
")
print(cycleway_stats)

# Longest roads by name
longest_roads = dbGetQuery(con, "
  SELECT 
    name,
    COUNT(*) as segments,
    SUM(ST_Length(geom)) as total_length_m
  FROM leeds_network
  WHERE name IS NOT NULL
  GROUP BY name
  ORDER BY total_length_m DESC
  LIMIT 10
")
print(longest_roads)
```

### Python

```{python}
#| label: duckdb-network-py
#| eval: false
# Total cycleway length
cycleway_stats = con.execute("""
  SELECT 
    highway,
    COUNT(*) as n_segments,
    SUM(ST_Length(geom)) as total_length_m,
    AVG(ST_Length(geom)) as avg_segment_length_m
  FROM leeds_network
  WHERE highway IN ('cycleway', 'path', 'footway')
  GROUP BY highway
  ORDER BY total_length_m DESC
""").df()
print(cycleway_stats)

# Longest roads by name
longest_roads = con.execute("""
  SELECT 
    name,
    COUNT(*) as segments,
    SUM(ST_Length(geom)) as total_length_m
  FROM leeds_network
  WHERE name IS NOT NULL
  GROUP BY name
  ORDER BY total_length_m DESC
  LIMIT 10
""").df()
print(longest_roads)
```

:::

## Part 4: Reflection & Discussion (15 mins)

### Questions to Discuss

1.  **When is DuckDB better than dplyr/pandas?** Think about data size, query complexity, and workflow.
2.  **How did AI help you learn?** Was the AI-generated code always correct? Did you need to modify it?
3.  **What surprised you?** About DuckDB? About using AI to learn?

### The AI-Assisted Learning Workflow

The approach we used in this session:

1.  **Ask** AI to explain a concept (DuckDB basics)
2.  **Try** writing code yourself based on the explanation
3.  **Compare** your code with AI's suggestion
4.  **Question** anything you don't understand — ask follow-up prompts
5.  **Apply** the concept to your own transport problem

This workflow is more powerful than copying AI output because you **understand** what the code does.

## Key Takeaways

-   DuckDB is a fast, embedded database ideal for transport data analysis
-   SQL skills transfer across R, Python, and many other tools
-   AI can accelerate learning new tools, but you must engage critically
-   Always understand code before using it in production

## Next Steps

-   [Day 2: Effective Prompting](s7.qmd) — Advanced AI interaction techniques
-   [Bonus: DuckDB challenges](bonus.qmd#duckdb-challenges) — More complex queries for experienced users
