---
title: "Session 6: AI-Driven Data Science for Transport Planning"
execute:
    eval: false
---

## Introduction

This session demonstrates how AI can accelerate **data science workflows** for transport planning. We use Leeds data throughout and show how the same analytical tasks can be accomplished with different tools — the AI adapts to your preferred framework.

::: callout-important
## Upskilling, not de-skilling
We will use AI to explain concepts and generate starter code, but you must **read, understand, and question** every output. If you don't understand what the code does, ask the AI to explain it. The aim is to leave this session *more capable* than when you started.
:::

## Choosing Your Tools

Transport data science can be done with many tool combinations. Here are the main options:

| Approach | R | Python | SQL |
|----------|---|--------|-----|
| **Base/low-level** | base R | Polars | DuckDB |
| **Tidyverse-style** | dplyr, tidyr | Pandas | DuckDB |
| **Spatial** | sf, terra | geopandas | DuckDB + spatial ext. |

**DuckDB** is notable because it:

-   Works as an **in-process SQL database** (no server needed)
-   Runs inside both R and Python
-   Provides a **language-agnostic** query interface — the same SQL works everywhere
-   Delivers high performance on large analytical datasets
-   Supports spatial data via its `spatial` extension

::: callout-tip
## Why we use DuckDB in examples
DuckDB provides a common language (SQL) that works across R and Python. By specifying DuckDB in your prompts, you get portable, high-performance code. Without this specification, AI may default to older frameworks like Pandas (which works but is slower for large data) or suggest approaches that do not translate between languages.
:::

**You can do everything in this session with other tools.** Ask the AI to translate any example to:

-   **R tidyverse** (dplyr + sf) — great for readable, pipe-based workflows
-   **Python Polars** — modern, fast alternative to Pandas
-   **Python Pandas** — widely used, though slower on large data
-   **Base R** — no dependencies, works everywhere

## Part 1: Loading and Exploring Data (30 mins)

### Step 1: Ask AI to Write Data Loading Code

Try this prompt, first without specifying a tool, then with one:

**Without tool specification:**
```
Write code to load leeds_network.geojson and count road segments by highway type
```

**With tool specification:**
```
Write R code using DuckDB to load leeds_network.geojson and count road 
segments by highway type, sorted by count descending. Explain each step.
```

::: {.panel-tabset}

### R + DuckDB

```{r}
#| label: load-duckdb-r
#| eval: false
library(duckdb)
library(sf)

# Connect to DuckDB (in-memory)
con = dbConnect(duckdb())

# Install and load spatial extension
dbExecute(con, "INSTALL spatial")
dbExecute(con, "LOAD spatial")

# Load the Leeds network data
dbExecute(con, "
  CREATE TABLE leeds_network AS 
  SELECT * FROM st_read('leeds_network.geojson')
")

# Query: count segments by highway type
result = dbGetQuery(con, "
  SELECT highway, COUNT(*) as n_segments
  FROM leeds_network
  GROUP BY highway
  ORDER BY n_segments DESC
")
print(result)
```

### Python + DuckDB

```{python}
#| label: load-duckdb-py
#| eval: false
import duckdb

# Connect to DuckDB (in-memory)
con = duckdb.connect()

# Install and load spatial extension
con.execute("INSTALL spatial")
con.execute("LOAD spatial")

# Load the Leeds network data
con.execute("""
  CREATE TABLE leeds_network AS 
  SELECT * FROM st_read('leeds_network.geojson')
""")

# Query: count segments by highway type
result = con.execute("""
  SELECT highway, COUNT(*) as n_segments
  FROM leeds_network
  GROUP BY highway
  ORDER BY n_segments DESC
""").df()
print(result)
```

### R + tidyverse

```{r}
#| label: load-tidyverse-r
#| eval: false
library(sf)
library(dplyr)

# Load the Leeds network data
leeds_network = read_sf("leeds_network.geojson")

# Count segments by highway type
result = leeds_network |>
  st_drop_geometry() |>
  count(highway, sort = TRUE)
print(result)
```

### Python + Polars

```{python}
#| label: load-polars-py
#| eval: false
import polars as pl
import geopandas as gpd

# Load the Leeds network data
leeds_network = gpd.read_file("leeds_network.geojson")

# Count segments by highway type
result = (
    pl.from_pandas(leeds_network.drop(columns="geom"))
    .group_by("highway")
    .len()
    .sort("len", descending=True)
)
print(result)
```

:::

::: {.callout-tip}
## Notice
All four approaches produce the same result. The SQL in DuckDB is identical for R and Python — this is the interoperability advantage. The tidyverse and Polars code is more idiomatic for each language but does not translate directly.
:::

### Step 2: Explore with AI

Use Copilot Chat to explore the data:

1.  `What columns are in leeds_network.geojson and what do they mean?`
2.  `Write [DuckDB/tidyverse/Polars] code to calculate total network length by highway type`
3.  `Create a bar chart showing the number of features per highway type`

## Part 2: OD Data Analysis (45 mins)

Now we bring in origin-destination data. The R `od` package provides example datasets.

### Step 3: Loading and Analysing OD Data

Ask Copilot:

```
The R 'od' package has example OD datasets for Leeds. Write code using DuckDB to:
1. Load the od::od_leeds dataset (export to CSV first)
2. Query the top 10 OD pairs by bicycle count
3. Calculate total trips by mode

Show both R and Python versions.
```

::: {.panel-tabset}

### R + DuckDB

```{r}
#| label: od-duckdb-r
#| eval: false
library(duckdb)
library(od)

# Get Leeds OD data and save for DuckDB
od_data = od::od_leeds
write.csv(od_data, "leeds_od_data.csv", row.names = FALSE)

# Connect and query
con = dbConnect(duckdb())
dbExecute(con, "LOAD spatial")

dbExecute(con, "
  CREATE TABLE od_data AS 
  SELECT * FROM read_csv_auto('leeds_od_data.csv')
")

# Top 10 cycling flows
top_cycle = dbGetQuery(con, "
  SELECT * FROM od_data
  ORDER BY bicycle DESC
  LIMIT 10
")
print(top_cycle)

# Mode totals
mode_totals = dbGetQuery(con, "
  SELECT 
    SUM(bicycle) as total_bicycle,
    SUM(walk) as total_walk,
    SUM(car_driver) as total_car,
    SUM(all) as total_all
  FROM od_data
")
print(mode_totals)
```

### R + tidyverse

```{r}
#| label: od-tidyverse-r
#| eval: false
library(od)
library(dplyr)

od_data = od::od_leeds

# Top 10 cycling flows
top_cycle = od_data |>
  arrange(desc(bicycle)) |>
  slice_head(n = 10)
print(top_cycle)

# Mode totals
mode_totals = od_data |>
  summarise(
    total_bicycle = sum(bicycle),
    total_walk = sum(walk),
    total_car = sum(car_driver),
    total_all = sum(all)
  )
print(mode_totals)
```

:::

### Step 4: Why Specify Your Tool?

Try these two prompts and compare the outputs:

**Vague prompt:**
```
Analyse the Leeds OD data and show cycling patterns
```

**Specific prompt:**
```
Using DuckDB in R, load leeds_od_data.csv and calculate cycling mode share 
(bicycle / all * 100) for each OD pair. Filter to flows with more than 50 
trips and sort by cycling mode share descending. Show the top 20.
```

The specific prompt will produce working code. The vague prompt may produce anything.

::: callout-tip
## Prompting tip
Specifying "DuckDB" in your prompt tells the AI to use modern, portable SQL. Without it, you might get Pandas code (Python) which works but is slower on large data, or base R code which may not scale. You can also specify "tidyverse" or "Polars" to get idiomatic code for your preferred framework.
:::

## Part 3: Spatial Analysis (30 mins)

### Step 5: Network Analysis

Ask Copilot:

```
Using [DuckDB / tidyverse / Polars], write code to:
1. Calculate the total length of cycleway in metres from leeds_network.geojson
2. Find the average length of road segments by highway type
3. Identify the 10 longest roads by name

Explain the approach used.
```

::: {.panel-tabset}

### R + DuckDB

```{r}
#| label: spatial-duckdb-r
#| eval: false
# Total cycleway length and stats
cycleway_stats = dbGetQuery(con, "
  SELECT 
    highway,
    COUNT(*) as n_segments,
    ROUND(SUM(ST_Length(geom))) as total_length_m,
    ROUND(AVG(ST_Length(geom))) as avg_segment_length_m
  FROM leeds_network
  WHERE highway IN ('cycleway', 'path', 'footway')
  GROUP BY highway
  ORDER BY total_length_m DESC
")
print(cycleway_stats)

# Longest roads by name
longest_roads = dbGetQuery(con, "
  SELECT 
    name,
    COUNT(*) as segments,
    ROUND(SUM(ST_Length(geom))) as total_length_m
  FROM leeds_network
  WHERE name IS NOT NULL
  GROUP BY name
  ORDER BY total_length_m DESC
  LIMIT 10
")
print(longest_roads)
```

### R + tidyverse

```{r}
#| label: spatial-tidyverse-r
#| eval: false
library(sf)
library(dplyr)

leeds_network = read_sf("leeds_network.geojson")

# Cycleway stats
cycleway_stats = leeds_network |>
  filter(highway %in% c("cycleway", "path", "footway")) |>
  mutate(length_m = as.numeric(st_length(geometry))) |>
  st_drop_geometry() |>
  group_by(highway) |>
  summarise(
    n_segments = n(),
    total_length_m = round(sum(length_m)),
    avg_segment_length_m = round(mean(length_m))
  ) |>
  arrange(desc(total_length_m))
print(cycleway_stats)

# Longest roads
longest_roads = leeds_network |>
  filter(!is.na(name)) |>
  mutate(length_m = as.numeric(st_length(geometry))) |>
  st_drop_geometry() |>
  group_by(name) |>
  summarise(
    segments = n(),
    total_length_m = round(sum(length_m))
  ) |>
  arrange(desc(total_length_m)) |>
  slice_head(n = 10)
print(longest_roads)
```

:::

## Part 4: Reflection & Discussion (15 mins)

### Questions to Discuss

1.  **Which tool fits your workflow?** DuckDB for interoperability? tidyverse/Polars for idiomatic code?
2.  **How did specifying the tool affect AI output?** Compare the quality of code when you do vs do not mention your preferred framework.
3.  **When would you switch tools?** At what data size does DuckDB become worth the SQL overhead?

### The AI-Assisted Learning Workflow

1.  **Specify your tool** — tell AI which framework you want
2.  **Ask** AI to explain the approach
3.  **Try** writing code yourself
4.  **Compare** with AI's suggestion
5.  **Question** anything unclear
6.  **Apply** to your own problems

## Key Takeaways

-   **Same analysis, many tools** — tidyverse, DuckDB, Polars all accomplish the same tasks
-   **DuckDB** provides language-agnostic SQL that works in both R and Python
-   **Specifying your tool** in prompts dramatically improves AI output quality
-   **AI can translate** between frameworks — ask it to convert tidyverse to Polars or vice versa
-   Always understand code before using it in production

## Next Steps

-   [Session 7: Applied Projects](s7.qmd) — Apply these tools to your own project
-   [Bonus: DuckDB challenges](bonus.qmd#duckdb-challenges) — More complex queries for experienced users
